Designing Types for R, Empirically

### 2020-05-16 paper/main.tex
corpus

For this paper we have selected \CorpusLoadable packages consisting of
\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native code
(C/C++/Fortran). Figure~\ref{fig:corpus} shows these packages, the size of the
dots reflects the size of the project in terms of lines of code including both R
and native code\footnote{ Lines of source code reported excludes comments and
  blank lines, counted by \emph{cloc}, \cf
  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the expression
code coverage in percents and the y-axis gives the number of reverse
dependencies in log scale. Dotted lines indicate means. Packages with over
\PackageSizeOutierRnd lines of code are annotated.

\begin{figure}[!h]  \centering
  \includegraphics[width=.8\linewidth]{plots/corpus.pdf}
  \caption{Corpus overview}\label{fig:corpus}
\end{figure}

@@

The \CorpusLoadable selected packages contain \CorpusRunnableCodeRnd lines of
runnable code in examples (\CorpusExamplesCodeRnd), tests (\CorpusTestsCodeRnd)
and vignettes (\CorpusVignettesCodeRnd). Running this code results on average in
\CorpusMeanExprCoverage package code coverage (CRAN average is
\AllMeanExprCoverage). Together, there is \CorpusRevdesRnd (on average
\CorpusMeanRevdes, median \CorpusMedianRevdes; CRAN average is \AllMeanRevdes,
median \AllMedianRevdes) reverse dependencies with \CorpusRevdepRunnableCodeRnd
runnable lines of code resulting in \CorpusRevdepMeanCodeCoverage coverage (on
average) of the corpus packages.

Together there are \CorpusFunctionsRnd defined R functions
(\CorpusPublicFunctionsRnd are from packages' public API).
\CorpusSThreeFunctionsRnd are S3 functions. Packages in the corpus define
\CorpusSThreeClasses S3 classes.

The code coverage was computed for each package using
\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R
tool.  The reverse package dependencies were extracted from the package
metadata using builtin function.

\paragraph{Type usage}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Table~\ref{ctb} reports on the 10 most frequent types
occuring in the corpus. Scalar doubles account for 11\% function arguments,
whereas logicals represent 27\% of observed values.  Together, \lgl and \chr
account for 60\%.

\begin{table}[!h]
% BEGIN Autogenerated
\small
\begin{tabular}{lrrrr}
\toprule
Type & Args & \% of Args & Observations & \% of Obs.\\
\midrule
\dbl & 13,255 &     11.0 & 41,560,797 & 1.5\\
\lgl & 10,142 &      8.4 & 712,895,959 & 27.2\\
\null & 9,175 &      7.6 & 122,202,183 & 4.6\\
\chr & 9,171 &       7.5 & 794,058,588 & 30.3\\
\VEC\dbl & 7,811 &   6.5 & 29,272,232 & 1.1\\
\VARGS & 7,759 &     6.4 & 76,369,333 & 2.9\\
\any & 6,243 &       5.2 & 255,539,497 & 9.0\\
\VEC\chr & 4,576 &   3.8 & 88,233,523 & 3.3\\
\CLASS{matrix} & 4,522     & 3.7 & 32,452,555 & 1.2\\
\CLASS{data.frame} & 2,781 & 2.3 & 7,749,094 & 0.3\\
\bottomrule
\end{tabular}

% END Autogenerated
\caption{Top types of arguments in R}\label{ctb}
\end{table}

@@

In order to count the occurrences of each type, we subdivided them into {\it
  kinds}: For our purposes, the {\it kind} of a type is equivalent to it's
top-level type constructor, so all classes have kind ``class'', all lists
have kind ``list'', and so on.  We found that classes are the most common
kind of type, accounting for roughly 31\% of types.  In order, the most
common classes are matrices (12\%), data.frames (7.5\%), formulas (2\%),
factors (2\%), and tibbles (2\%).  Roughly 25\% of classes are part of R's
base language implementation, the others being defined as part of some
package's functionality.

Besides classes, scalars and vectors are the next most common kind, making
up 41\% of types. with scalars making up 28.1\% of types and vectors
12.65\%.  Nulls and lists follow at 8\% and 7\% respectively, and the vararg
type \ldots makes up 6.5\% of arguments.  This all totals up to over 90\% of
types.  We will discuss the any type in more detail in
Section~\ref{subsec:any}.

>>>>

For this paper we have selected \CorpusLoadable packages consisting of
\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native
code (C/Fortran). Figure~\ref{fig:corpus} shows these packages, the size of
the dots reflects the project's size in lines of code including both R and
native code\footnote{Lines of source code reported excludes comments and
  blank lines, counted by \emph{cloc}, \cf
  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
expression code coverage in percents and the y-axis gives the number of
reverse dependencies in log scale. Dotted lines indicate means. Packages
with over \PackageSizeOutierRnd lines of code are annotated.

\begin{figure}[!h]\centering\includegraphics[width=.8\linewidth]
  {plots/corpus.pdf}\caption{Corpus}\label{fig:corpus}
\end{figure}

@@

The \CorpusLoadable selected packages contain \CorpusRunnableCodeRnd lines of
runnable code in examples (\CorpusExamplesCodeRnd), tests (\CorpusTestsCodeRnd)
and vignettes (\CorpusVignettesCodeRnd). Running this code results on average in
\CorpusMeanExprCoverage package code coverage (CRAN average is
\AllMeanExprCoverage). Together, there is \CorpusRevdesRnd (on average
\CorpusMeanRevdes, median \CorpusMedianRevdes; CRAN average is \AllMeanRevdes,
median \AllMedianRevdes) reverse dependencies with \CorpusRevdepRunnableCodeRnd
runnable lines of code resulting in \CorpusRevdepMeanCodeCoverage coverage (on
average) of the corpus packages.
Together there are \CorpusFunctionsRnd defined R functions
(\CorpusPublicFunctionsRnd are from packages' public API).
\CorpusSThreeFunctionsRnd are S3 functions. Packages in the corpus define
\CorpusSThreeClasses S3 classes.
The code coverage was computed for each package using
\covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R
tool.  The reverse package dependencies were extracted from the package
metadata using builtin function.

\paragraph{Type usage}%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
During execution 3,147 different types were observed.  Classes are the most
common types, accounting for roughly 31\% of types of arguments.  The most
common classes are \c{matrices} (12\%), \c{data.frames} (7.5\%),
\c{formulas} (2\%), \c{factors} (2\%), and \c{tibbles} (2\%).  Roughly 25\%
of classes are part of R's base libraries, the others are user-defined.
Scalars and vectors are the next most common kind, making up 41\% of
remaining types. with scalars making up 28\% of types and vectors 12\%.
Nulls and lists follow at 8\% and 7\% respectively, and the vararg type
makes up 6\% of arguments.  This all totals up to over 90\% of types.
Table~\ref{ctb} reports on the 10 most frequent types occuring in the
corpus.

\begin{table}[!h]
% BEGIN Autogenerated
\small\begin{tabular}{lrrrr}\toprule
Type & Args & \% of Args & Observations & \% of Obs.\\\midrule
\dbl               & 13,255 &     11.0 &  41,560,797 & 1.5\\
\lgl               & 10,142 &      8.4 & 712,895,959 & 27.2\\
\null              &  9,175 &      7.6 & 122,202,183 & 4.6\\
\chr               &  9,171 &      7.5 & 794,058,588 & 30.3\\
\VEC\dbl           &  7,811 &      6.5 &  29,272,232 & 1.1\\
\VARGS             &  7,759 &      6.4 &  76,369,333 & 2.9\\
\any               &  6,243 &      5.2 & 255,539,497 & 9.0\\
\VEC\chr           &  4,576 &      3.8 &  88,233,523 & 3.3\\
\CLASS{\texttt{matrix}}     & 4,522 & 3.7 & 32,452,555 & 1.2\\
\CLASS{\texttt{data.fram}e} & 2,781 & 2.3 &  7,749,094 & 0.3\\
\bottomrule
\end{tabular}
\caption{Top types of arguments in R}\label{ctb}
\end{table}

@@
